Analyzing the Effect of Surface in Tennis Grand Slams

### 2018-09-15 contest/surface-type/paper-tennis.Rmd
updates

## Nadal

@@

## Federer
```{r federer-model}

@@

## Williams
```{r williams-model}

@@

>>>>

### Nadal

@@

### Federer
```{r federer-model}

@@

### Williams
```{r williams-model}

@@

## Mixed-effects models for all grand slams

### Logistic regression for wins

We will use a multilevel logistic regression framework. This allows us to include fixed effects, which remain the same for each player, as well as include player-level effects. Since players appear multiple times in the data, the observations are not independent and including a player-level effect allows us to account for this dependence. It also provides a way to examine individual player effects and assess, for instance, whether Federer is more likely to win at Wimbledon after accounting for other variables.

We initially planned to use points won by an individual as our outcome variable, but found that it may not be an ideal measure of player performance. For instance, players may score few points in a match due to a poor performance, but they may also score few (relative) points if they win a short match. Modeling the number of points won is also complicated by the difference between length of matches for men (best of 5 sets) and women (best of 3 sets). For these reasons, we chose to model whether a player won the match, with the following predictors:

* `ioc_fac`: Country that the player represents
* `tournament`: Which Grand Slam the match was played at
* `late_round`: Indicator for whether the match occured in Round of 16 or later
* `rank`: Rank of player at the time of the match
    + Included on log scale 
* `opponent_rank`: Rank of opponent at the time of the match
    + Included on log scale
* `year`: Factor variable with a level for each year included in the dataset (2013-2017)
* `atp`: Indicator that the match was played in the ATP league instead of the WTF league.

(short discussion of models fit, show plots for final model)

### Linear regression for other stats

Rank and opponent's rank clearly have the largest effect on predicting the winner. Looking at the individual effects for Serena, Nadal, and Federer; they actually fall near the average for all four grand slams. Including rank as an indicator (seeded/not-seeded) substantially decreases the fit of the model, even when individual effects are included. We now look at a similar model, but instead of using `win' as the outcome variable, we'll look at game statistics that may be more sensitive to court surface -- aces, winners, and net points won -- and see if individual differences are detectable.

(some plots that show a) Serena/Federer are stronger servers, b) Federer makes fewer unforced errors at Wimbledon, c) something about Nadal)
